Development of restored dike grasslands to reference needs more than four years and can hardly modified by local site conditions

### 2023-02-10 manuscript.Rmd
Update manuscript.Rmd

---
title: "Development of restored dike grasslands to reference needs more than four years and can hardly modified by local site conditions"
author: Markus Bauer*, Jakob Huber, Johannes Kollmann
output:
  officedown::rdocx_document:
    reference_docx: StylesTemplate.docx
csl: template-journal-of-applied-ecology.csl
bibliography: bibliography.bib
---

@@

Sowing as an approach to restore semi-natural grasslands has long been established [@Kiehl.2010], is also able to restore functional diversity [@Engst.2016] and is continuously developed [e.g., @Freitag.2021]. Restoration is seen as a key factor to sustain ecosystem services and biodiversity [@COP.2014; @UnitedNations.2019b]. It is necessary to successfully restore grasslands since this habitat type provide several ecosystem services and a high biodiversity while it is worldwide endangered [@Dengler.2014; @Bardgett.2021] and even by restoration through unsuited afforestation [@dudley2020]. In Europe, calcareous grasslands and hay meadows are on the red list [Category 3, vulnerable, @Janssen.2016].

Sowing high-diversity seed mixtures of local provenance produced by commercial but specialised seed companies is a promising way to scale up restoration efforts. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021]. For this reason, restoration ecology has to increase predictability of restoration approaches [@Suding.2011; @Mouquet.2015]. Rigorous, repeatable, and transparent experiments linked to a hypothesis derived from theory are needed to guide evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration approach are key predictors of the development of vegetation after sowing [@Brudvig.2017b]. Seeding in general is a way to overcome the dispersal filter [@kiehl2006], but we focus on environmental and the biotic filtering which are the main assembly processes to be manipulated by the choice of the right substrate-seed mixture combination [@Torok.2017]. This means that a mutually adaption of the substrate to the preferences of the target vegetation, and the seed mixtures to the characteristics of the chosen substrate. The right substrates should reduce habitat filtering and the right seed mixture should minimize competitive exclusion. Modifying seed mixtures for certain site conditions came up with functional plant traits [@Funk.2008; @Laughlin.2014] but it is not easy to put it into practice for novel ecosystems like urban grasslands [@Bauer.2022]. Substrates cannot only be modified for conservation purposes in urban areas, but also in quarries [@chenot-lescure2022] or dikes [@Liebrand.1996].

Dikes are promising sites for the restoration of species-rich grasslands [@Liebrand.1996], because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in an agricultural landscape [@Batori.2020]. Dikes have specific characteristics with strongly exposed slopes and river and land orientation [@Batori.2016] and need a reliable vegetation cover to conserve the functionality of the construction [@KleberLerchbaumer.2017]. Dikes can reconcile several requirements [@teixeira2022], but the substrate-seed mixture combination must be adapted.

@@

>>>>

---
title: "Development of restored dike grasslands to reference needs more than four years and can hardly modified by local site conditions"
author: Markus Bauer*, Jakob Huber, Johannes Kollmann
output:
  officedown::rdocx_document:
    reference_docx: template_word.docx
csl: template_references_j_appl_ecol.csl
bibliography: bibliography.bib
---

@@

Sowing as an approach to restore semi-natural grasslands has long been established [@Kiehl.2010], is also able to restore functional diversity [@Engst.2016] and is continuously developed [e.g., @Freitag.2021]. Restoration is seen as a key factor to sustain ecosystem services and biodiversity [@COP.2014; @UnitedNations.2019b]. It is necessary to successfully restore grasslands since this habitat type provide several ecosystem services and a high biodiversity while it is worldwide endangered [@Dengler.2014; @Bardgett.2021] even by restoration through unsuited afforestation [@dudley2020]. In Europe, calcareous grasslands and hay meadows are on the red list [Category 3, vulnerable, @Janssen.2016].

Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021]. For this reason, restoration ecology has to increase predictability of restoration approaches [@Suding.2011; @Mouquet.2015]. Rigorous, repeatable, and transparent experiments linked to a hypothesis derived from theory are needed to guide evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration approach are key predictors of the development of vegetation after sowing [@Brudvig.2017b]. Seeding in general is a way to overcome the dispersal filter [@kiehl2006; @orrock2023], but we focus on environmental and the biotic filtering which are the main assembly processes to be manipulated by the choice of the right seed mixture-substrate combination [@Torok.2017]. This means a mutually adaption of the substrate to the preferences of the target vegetation, and the seed mixtures to the characteristics of the chosen substrate. The right substrates should reduce habitat filtering and the right seed mixture should minimize competitive exclusion of unsuited species. Modifying seed mixtures for certain site conditions came up with functional plant traits [@Funk.2008; @Laughlin.2014] but it is not easy to put it into practice for novel ecosystems like urban grasslands [@Bauer.2022]. Substrates cannot only be modified for conservation purposes in urban areas, but also in quarries [@chenot-lescure2022] or dikes [@Liebrand.1996].

Dikes are promising sites for the restoration of species-rich grasslands [@Liebrand.1996], because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in an agricultural landscape [@Batori.2020]. Dikes have specific characteristics with strongly exposed slopes and river and land orientation [@Batori.2016] and need a reliable vegetation cover to conserve dike security [@KleberLerchbaumer.2017]. Dikes can reconcile several requirements [@teixeira2022], but the seed mixture-substrate combination must be adapted.

@@

The vegetation of the dike experiment developed in the right direction, but with reduced velocity and significantly and lasting worse development of the south exposition. Sand admixture of 25% was sufficient to reach a positive effect. All in all, it was hardly possible to identify an ideal seed mixture-substrate combination. Though, hay meadows developed closer to its reference on the north exposition than dry grasslands.

## The effect of seed mixture-substrate combinations

Persistence, FCS

sand admixture positive but 25% enough (ref. to chenot?)

substrate depth does not matter (what can be discussed?)

Higher seeding density not necessary for biodiversity (ref to practioners, evtl. to Batori?)

no significant effect of combination (variability important, further research)

hay meadows closer to reference

Lasting effect of south exposition (drought, seeding late)

## The success of the restoration approaches
